---
title: "Marginal Analysis Example"
author: Kris Sankaran
output: readthedown::html_document
---

```{r}
library(MIGsim) # devtools::install_github("krisrs1128/intro-to-simulation/MIGsim")
library(SummarizedExperiment)
library(gamboostLSS)
library(ggdist)
library(glue)
library(splines)
library(tidyverse)
library(scDesigner)
```

### Using SummarizedExperiment

* How many genera are available in this experiment object?
* What was the most common phylum in this dataset?
* What was the average participant age?

```{r}
data(atlas)
dim(atlas) # 125 genera
table(rowData(atlas)$Phylum)
mean(colData(atlas)$age)

# or just,
mean(atlas$age)
```

### Demo: Gaussian over Time

```{r}
data(exper_ts)
exper_lineplot(exper_ts)
```

```{r}
sim <- setup_simulator(exper_ts, ~ ns(time, df = 7) * group, ~ GaussianLSS()) |>
  estimate(nu = 0.01, mstop = 1000)

sample(sim) |>
  exper_lineplot()
```

```{r}
sim <- sim |>
  mutate(4:6, link = ~ ns(time, df = 7)) |>
  estimate(nu = 0.01, mstop = 1000)

sample(sim) |>
  exper_lineplot()
```


### Exercise: Gaussian Mixture

```{r}
data(exper_grouped)
exper_histogram(exper_grouped)
```

* Explain how the simulation code below works. 
* Modify it so that it accurate reflects the difference between groups for the first three genes.

```{r}
sim <- setup_simulator(exper_grouped, ~1, ~ GaussianLSS()) |>
  estimate()

sample(sim) |>
  exper_histogram()
```

```{r}
sim <- sim |>
  mutate(1:3, link = ~group) |>
  estimate()

sample(sim) |>
  exper_histogram()
```

### Estimate and Visualize

```{r}
data(atlas)

fmla <- list(
  mu = ~ bmi_group + log_depth,
  sigma = ~ bmi_group + log_depth,
  nu = ~1
)
sim <- setup_simulator(atlas, fmla, ~ ZINBLSS()) |>
  estimate(nu = 0.005, mstop = 1000)
```

Visualizing simulator quality.

```{r}
atlas_ <- sample(sim)
combined <- bind_rows(
  real = pivot_experiment(atlas), # real data
  simulated = pivot_experiment(atlas_), # simulated
  .id = "source"
)

ggplot(combined) +
  geom_boxplot(
    aes(asinh(value), reorder(feature, value, median), fill = bmi_group)
  ) +
  facet_grid(. ~ source)
```

### Apply to Power Analysis

Re-estimate the model with all taxa initially set to NULL.

```{r}
null_fmla <- list(mu = ~log_depth, sigma = ~log_depth, nu = ~1)
```

```{r}
nulls <- run_DA_method(atlas, "LIMMA_voom") |>
  rownames() |>
  tail(16)

sim <- sim |>
  mutate(dplyr::any_of(nulls), link = null_fmla) |>
  estimate(nu = 0.005, mstop = 500)
```

```{r}
config <- expand.grid(
  sample_size = floor(seq(50, 1200, length.out = 5)),
  n_rep = 1:10
) |>
  mutate(run = row_number())

results <- list()
for (i in seq_len(nrow(config))) {
  atlas_ <- sample_n(sim_, config$sample_size[i])
  results[[i]] <- run_DA_method(atlas_, "LIMMA_voom") |>
    da_metrics(nonnull, level = 0.3)
  print(glue("{i}/{nrow(config)}"))
}
```

```{r}
results <- bind_rows(results, .id = "run") |>
  left_join(config) |>
  ggplot() +
  stat_pointinterval(aes(
    factor(sample_size), power
  )) +
  facet_wrap(~ metric, scales = "free")
```
